---
title: "Brazil: Unemployment"
categories: ['data-visualization', 'english', 'ggplot2']
date: "2023-10-20"
execute:
  warning: false
  message: false
  echo: false
draft: true
---

```{r, setup}
knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%",
  fig.width = 9
)
```

```{r, libs}
library(sidrar)
library(ggplot2)
library(dplyr)
library(tidyr)
library(showtext)
library(sf)
library(ggrepel)

font_add_google("Merriweather", "Merriweather")
font_add_google("Roboto Condensed", "Roboto Condensed")
font_add_google("Roboto Condensed", "Roboto Condensed")

showtext_auto()
```

```{r, import-data}
state <- geobr::read_state(showProgress = FALSE)
dim_state <- as_tibble(sf::st_drop_geometry(state))
unemp <- get_sidra(
  4099,
  variable = 4099,
  geo = "State",
  period = "201201-202302"
)
```

```{r, clean-data}
#> Clean Sidra table
tbl_unemp <- unemp |>
  janitor::clean_names() |>
  select(
    code_state = unidade_da_federacao_codigo,
    trimestre = trimestre_codigo,
    unemp = valor
  ) |>
  mutate(
    date = zoo::as.Date(zoo::as.yearqtr(trimestre, format = "%Y%q")),
    code_state = as.numeric(code_state)
  ) |>
  left_join(dim_state, by = "code_state")

#> Lollipop data
lvls <- tbl_unemp |>
  filter(date == as.Date("2019-04-01")) |>
  arrange(unemp) |>
  pull(abbrev_state)

tbl_compare <- tbl_unemp |>
  filter(date %in% c(as.Date("2019-04-01"), as.Date("2023-04-01"))) |>
  mutate(abbrev_state = factor(abbrev_state, levels = lvls))

tbl_chg <- tbl_compare |>
  mutate(chg = unemp - lag(unemp), .by = "code_state") |>
  filter(!is.na(chg)) |>
  select(code_state, chg, unemp)


#> Map data
map_chg <- left_join(state, tbl_chg, by = "code_state")

map_chg <- map_chg %>%
  mutate(
    x = as.numeric(st_coordinates(st_centroid(.))[, "X"]),
    y = as.numeric(st_coordinates(st_centroid(.))[, "Y"]),
    label = round(unemp, 2),
    repel = if_else(code_state %in% c(24:28, 53), 1L, 0L),
    group = findInterval(chg, c(6, 10, 12, 16))
  )
```

```{r lollipop}
#| out-width: 80%
#| fig-width: 8
#| fig-height: 9
ggplot(tbl_compare, aes(x = abbrev_state, y = unemp)) +
  geom_line(aes(group = abbrev_state), color = "gray40") +
  geom_point(aes(color = as.factor(date)), size = 2) +
  scale_y_continuous(breaks = seq(0, 18, 2)) +
  coord_flip() +
  scale_color_manual(
    name = "",
    values = c("#3690c0", "#016c59"),
    labels = c("2019-Q2", "2023-Q2")
  ) +
  labs(
    title = "Unemployment rate fell in all Brazilian states",
    subtitle = "Comparison of 2019-Q2 unemployment rate with 2023-Q2. Includes informal workers.",
    caption = "Source: IBGE (PNADC)",
    x = NULL,
    y = "(%)"
  ) +
  theme_minimal(base_family = "Roboto Condensed") +
  theme(
    plot.title = element_text(family = "Merriweather", size = 16),
    plot.subtitle = element_text(
      family = "Merriweather",
      size = 11,
      color = "gray10"
    ),
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

```{r, map}
ggplot(map_chg) +
  geom_sf(aes(fill = unemp), linewidth = 0.3, color = "white") +
  geom_label(
    data = filter(map_chg, repel == 0),
    aes(x = x, y = y, label = label),
    size = 3,
    family = "Roboto Condensed",
    color = "gray10"
  ) +
  geom_label_repel(
    data = filter(map_chg, repel == 1),
    aes(x = x, y = y, label = label),
    size = 3,
    family = "Roboto Condensed",
    color = "gray10",
    nudge_x = 5
  ) +
  labs(
    title = "Southern and Mid-Western States\nHave Lowest Unemployment Rates"
  ) +
  scale_fill_distiller(name = "Unemployment (%)", palette = "PuBuGn") +
  ggthemes::theme_map() +
  theme(
    plot.title = element_text(family = "Merriweather", size = 16),
    legend.text = element_text(family = "Roboto Condensed"),
    legend.title = element_text(family = "Roboto Condensed")
  )
```

```{r}
labor_share_br <- get_sidra(
  6461,
  variable = 4096,
  period = "201801-202304",
  geo = "Brazil"
)

tbl_labor_share_br <- labor_share_br |>
  janitor::clean_names() |>
  select(trimestre = trimestre_codigo, share = valor) |>
  mutate(
    date = zoo::as.Date(zoo::as.yearqtr(trimestre, format = "%Y%q"))
  )

base_index <- tbl_labor_share_br |>
  filter(date >= as.Date("2019-01-01"), date <= as.Date("2019-12-01")) |>
  summarise(base = mean(share))

tbl_labor_share_br <- tbl_labor_share_br |>
  mutate(index = share / base_index$base * 100)
```

```{r}
labor_share <- get_sidra(
  6461,
  variable = 4096,
  period = "201801-202304",
  geo = "State"
)

tbl_labor_share <- labor_share |>
  janitor::clean_names() |>
  select(
    code_state = unidade_da_federacao_codigo,
    trimestre = trimestre_codigo,
    share = valor
  ) |>
  mutate(
    date = zoo::as.Date(zoo::as.yearqtr(trimestre, format = "%Y%q")),
    code_state = as.numeric(code_state)
  ) |>
  left_join(dim_state, by = "code_state")

index_base <- tbl_labor_share |>
  filter(date >= as.Date("2019-01-01"), date <= as.Date("2019-12-01")) |>
  summarise(base = mean(share), .by = "code_state")

tbl_labor_share <- tbl_labor_share |>
  left_join(index_base, by = "code_state") |>
  mutate(
    index = share / base * 100,
    highlight = if_else(
      abbrev_state %in% c("AL", "AC", "SP", "RJ", "TO", "RN"),
      1L,
      0L
    )
  )
```

```{r}
ggplot() +
  geom_line(
    data = tbl_labor_share,
    aes(
      x = date,
      y = index,
      group = code_state,
      color = as.factor(name_region)
    ),
    alpha = 0.7
  ) +
  geom_line(
    data = tbl_labor_share_br,
    aes(x = date, y = index),
    color = "#000000",
    linewidth = 1
  ) +
  geom_hline(yintercept = 100) +
  geom_text(
    data = filter(tbl_labor_share, date == max(date), highlight == 1),
    aes(x = date, y = index, label = abbrev_state),
    family = "Roboto Condensed",
    size = 3,
    nudge_x = 50
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_color_manual(
    name = "",
    values = c("#377eb8", "#ff7f00", "#a65628", "#4daf4a", "#984ea3")
  ) +
  labs(
    title = "Drop in labor participation rate",
    y = "Index (2019 = 100)",
    x = NULL,
    caption = "Source: IBGE (PNADC)"
  ) +
  theme_minimal(base_family = "Roboto Condensed") +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "top",
    plot.title = element_text(family = "Merriweather", size = 16)
  )
```

# Unemployment

Brazilian unemployment has fallen sharply since the end of the pandemic and is below pre-pandemic values. In this Chart of the Week I highlight the decrease in unemployment across all Brazilian states. The northern state of Roraíma registered the biggest drop in the past three years: the rate of unemployment fell from 15.1% to 5.1%.

It should be noted that the Brazilian labor market has a large degree of informality. This overall unemployment rate

```{r}
tbl_unemp |>
  filter(abbrev_state == "RR", date %in% c(as.Date("2023-04-01")))
```
