---
title: "Weekly Viz: Transportation in São Paulo"
date: "2023-11-03"
categories: ['data-visualization', 'weekly-viz', 'english', 'sao-paulo']
description: "This week, I delve into the transportation modes within the Greater São Paulo Region by analyzing data sourced from the Origin Destination Survey. The achievement of recently announced sustainability goals, targeting net-zero emissions by 2050, hinges significantly on transportation choices."
execute: 
  echo: false
  warning: false
  message: false
---

# São Paulo Transportation

The Metropolitan Region of São Paulo (MRSP) accommodates over 20 million residents and facilitates over 40 million trips on a daily basis. The city of São Paulo recently announced that it aims to drastically reduce greenhouse gas emissions, achieving net-zero emissions by 2050. To attain this objective, the city must focus on improving transit options and reducing car-dependency.

According to the [most recent Metro Report](https://transparencia.metrosp.com.br/sites/default/files/SÍNTESE_OD2017_ago19.pdf), published in 2018, travel patterns in the Metropolitan Region of São Paulo are categorized as follows: one-third of journeys involve non-motorized modes (such as walking or cycling), while two-thirds are conducted through motorized means. Within the latter category, 55% of trips utilize collective modes of transportation (bus, subway, train), and 45% rely on individual modes (car, taxi, motorcycle).

Notably, trips undertaken in private automobiles (excluding taxis) constitute around 27% of the total daily trips. In other words, roughly one in every four journeys in the metro region involves private car usage.

The diagram below delineates the entirety of daily trips within the MRSP. It's important to acknowledge that the data pertains to the year 2017, during which many stations along the 4-Yellow and 5-Coral lines were still under construction. Similarly, the 13-Jade and 15-Silver lines (monorail) were not yet fully operational. For the sake of simplicity, the analysis encompasses all types of travel, although commuting between home and work constitutes the predominant share of this set.

```{r}
library(plotly)
```

```{r}
fig <- plot_ly(
  type = "sankey",
  orientation = "h",
  
  node = list(
    label = c(
      "Motorized", "Shared", "Private", "Bus", "Chartered", "School",
      "Metro", "Train", "Automobile", "Taxi", "Motorcycle", "Non-motorized", "On foot",
      "Bicycle", "Trips"
      ),
    color = c("#08519c", "#2171b5", "#238b45", "#6baed6", "#9ecae1", "#c6dbef",
              "#6a51a3", "#9e9ac8", "#41ab5d", "#a1d99b", "#c7e9c0", "#fc4e2a",
              "#fd8d3c", "#feb24c"),
    pad = 15,
    thickness = 15,
    line = list(
      color = "black",
      width = 0.5
    )
  ),
  size = I(20),
  link = list(
    source = c(0,0,1,1,1,1,1,2,2,2,11,11),
    target = c(1,2,3,4,5,6,7,8,9,10,12,13),
    value =  c(36.4,30.9,19.8,0.61,4.98,8.1,3,27, 1.14, 2.56, 31.8, 0.9)
  )
)

fig <- fig %>% layout(
  title = list(
    text = "Travels in the Greater São Paulo",
    y = 0.95),
  font = list(
    size = 20,
    family = "Gill Sans"
  )
)

fig
```
