---
title: ""
date: "2025-07-15"
description: ""
categories: ["data-science", "tutorial-R"]
format: html
execute: 
  warning: false
  message: false
draft: true
---

# Instalando o R

Recentemente, ...

```{r}
library(pak)
```

## Pacotes

### Data Cleaning

```{r}

packs_core_data <- c(
    "tidyverse",
    "data.table",
    "janitor",
    "lubridate",
    "glue",
    "fs",
    "here",
    "dtplyr"
)

packs_import_data <- c(
    "foreign",
    "readxl",
    "haven",
    "jsonlite",
    "vroom",
    "qs",
    "arrow",
    "rio"
)

packs_database <- c(
    "RPostgres",
    "DBI",
    "odbc",
    "dbplyr",
    "RSQLite",         
    "bigrquery",       
    "googledrive",
    "sparklyr"        
)
```

### Data Visualization

```{r}
packs_data_viz <- c(
  # Core package (already should be installed by 'tidyverse')
  "ggplot2",         
  # ggplot2 extensions
  "patchwork",       # Combining ggplot2 plots
  "ggrepel",         # Avoid overlapping text labels
  "ggcorrplot",      # Correlation matrix visualization
  "gganimate",       # Animated plots
  "ggtext",          # Improved text rendering

  # Specialized visualizations
  "ggalluvial",      # Alluvial diagrams
  "ggbump",
  "ggridges",        # Ridge plots
  "ggdist",          # Visualizing distributions
  "GGally",

  # Interactive visualization
  "plotly",          # Interactive plots
  "DT",              # Interactive data tables
  "htmlwidgets",     # HTML widgets framework
  
  # Color palettes, custom fonts, and extra themes
  "ggthemes",
  "hrbrthemes",
  "showtext",
  "ragg",
  "viridis",         # Colorblind-friendly palettes
  "RColorBrewer",    # Color palettes for maps and charts
  "ggsci",           # Scientific journal color palettes
  "wesanderson"      # Wes Anderson movie color palettes
  "MetBrewer",
  "paletteer"
)
```

### Spatial
---

```{r}
packs_core_spatial <- c(
  "sf",
  "terra",
  "lwgeom",
  ""
)

packs_spatial_web <- c(
  "tidygeocoder",
  "tidygtfs",
  "osmdata",
  "osrm",
  "mapview"
)


  # Core spatial packages
  "sf",              # Simple features for vector data
  "terra",           # Spatial data analysis (successor to raster)
  "stars",           # Spatiotemporal arrays
  "lwgeom",          # Advanced geometric operations
  
  # Spatial statistics
  "spdep",           # Spatial dependence analysis
  "spatstat",        # Spatial point pattern analysis
  "gstat",           # Geostatistics
  "automap",         # Automated kriging
  
  # Geocoding and web services
  "tidygeocoder",    # Geocoding interface
  "osmdata",         # OpenStreetMap data
  "elevatr",         # Elevation data access
  
  # Spatial visualization
  "leaflet",         # Interactive maps
  "mapview",         # Interactive spatial data viewing
  "tmap",            # Thematic mapping
  "ggmap",           # ggplot2 extension for maps
  "rayshader",       # 3D mapping and visualization
  
  # Coordinate reference systems
  "rmapshaper",      # Spatial data simplification
  "units"            # Measurement units handling
```

## Configurando o RStudio



```{r}

```

# Essential R Packages for Data Science
# Author: [Your Name]
# Date: July 2025
# Description: Comprehensive package installation for spatial analysis, 
#              time series, data visualization, and machine learning

# Install pak if not already installed
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}

library(pak)

# =============================================================================
# CORE DATA MANIPULATION & GENERAL UTILITIES
# =============================================================================
cat("Installing core data manipulation packages...\n")
pak::pak(c(
  "tidyverse",        # Complete data science toolkit (dplyr, ggplot2, etc.)
  "data.table",       # Fast data manipulation
  "dtplyr",          # dplyr backend for data.table
  "janitor",         # Data cleaning utilities
  "here",            # Project-relative file paths
  "fs",              # File system operations
  "glue",            # String interpolation
  "lubridate",       # Date/time manipulation
  "readxl",          # Excel file reading
  "writexl",         # Excel file writing
  "haven",           # SPSS, Stata, SAS file import
  "jsonlite",        # JSON data handling
  "httr2"            # HTTP requests (successor to httr)
))

# =============================================================================
# SPATIAL ANALYSIS & GIS
# =============================================================================
cat("Installing spatial analysis packages...\n")
pak::pak(c(
  # Core spatial packages
  "sf",              # Simple features for vector data
  "terra",           # Spatial data analysis (successor to raster)
  "stars",           # Spatiotemporal arrays
  "lwgeom",          # Advanced geometric operations
  
  # Spatial statistics
  "spdep",           # Spatial dependence analysis
  "spatstat",        # Spatial point pattern analysis
  "gstat",           # Geostatistics
  "automap",         # Automated kriging
  
  # Geocoding and web services
  "tidygeocoder",    # Geocoding interface
  "osmdata",         # OpenStreetMap data
  "elevatr",         # Elevation data access
  
  # Spatial visualization
  "leaflet",         # Interactive maps
  "mapview",         # Interactive spatial data viewing
  "tmap",            # Thematic mapping
  "ggmap",           # ggplot2 extension for maps
  "rayshader",       # 3D mapping and visualization
  
  # Coordinate reference systems
  "rmapshaper",      # Spatial data simplification
  "units"            # Measurement units handling
))

# =============================================================================
# TIME SERIES ANALYSIS
# =============================================================================
cat("Installing time series analysis packages...\n")
pak::pak(c(
  # Core time series
  "forecast",        # Comprehensive forecasting methods
  "fable",           # Tidy time series forecasting
  "tsibble",         # Tidy time series data frames
  "feasts",          # Feature extraction and statistics
  "fabletools",      # Tools for time series modeling
  
  # Advanced time series methods
  "prophet",         # Facebook's forecasting procedure
  "bsts",            # Bayesian structural time series
  "vars",            # Vector autoregression
  "urca",            # Unit root and cointegration tests
  "rugarch",         # Univariate GARCH models
  
  # Time series decomposition
  "seasonal",        # X-13ARIMA-SEATS seasonal adjustment
  "x12",             # X-12-ARIMA seasonal adjustment
  "stl",             # Seasonal-trend decomposition
  
  # High-frequency time series
  "xts",             # Extensible time series
  "zoo",             # Z's ordered observations
  "quantmod",        # Quantitative financial modeling
  "TTR",             # Technical trading rules
  
  # Time series visualization
  "dygraphs",        # Interactive time series plots
  "plotly"           # Interactive plots (also useful for other viz)
))

# =============================================================================
# DATA VISUALIZATION
# =============================================================================
cat("Installing data visualization packages...\n")
pak::pak(c(
  # ggplot2 extensions
  "ggplot2",         # Grammar of graphics (included in tidyverse)
  "patchwork",       # Combining ggplot2 plots
  "ggrepel",         # Avoid overlapping text labels
  "ggridges",        # Ridge plots
  "ggdist",          # Visualizing distributions
  "ggcorrplot",      # Correlation matrix visualization
  "ggalluvial",      # Alluvial diagrams
  "ggraph",          # Network visualization with ggplot2
  "gganimate",       # Animated plots
  "ggtext",          # Improved text rendering
  
  # Specialized visualization
  "corrplot",        # Correlation matrix plots
  "pheatmap",        # Pretty heatmaps
  "ComplexHeatmap",  # Advanced heatmap visualization
  "circlize",        # Circular visualization
  "networkD3",       # D3 network visualizations
  "visNetwork",      # Network visualization
  "DiagrammeR",      # Graph and flowchart creation
  
  # Interactive visualization
  "plotly",          # Interactive plots
  "DT",              # Interactive data tables
  "crosstalk",       # Inter-widget communication
  "htmlwidgets",     # HTML widgets framework
  
  # Color palettes
  "viridis",         # Colorblind-friendly palettes
  "RColorBrewer",    # Color palettes for maps and charts
  "ggsci",           # Scientific journal color palettes
  "wesanderson"      # Wes Anderson movie color palettes
))

# =============================================================================
# MACHINE LEARNING & STATISTICAL MODELING
# =============================================================================
cat("Installing machine learning packages...\n")
pak::pak(c(
  # ML frameworks
  "tidymodels",      # Tidy modeling framework
  "caret",           # Classification and regression training
  "mlr3",            # Machine learning in R (next generation)
  "mlr3verse",       # Extended mlr3 ecosystem
  
  # Specific algorithms
  "randomForest",    # Random forests
  "ranger",          # Fast random forests
  "xgboost",         # Gradient boosting
  "lightgbm",        # Microsoft's gradient boosting
  "glmnet",          # Regularized linear models
  "e1071",           # SVM and other algorithms
  "kernlab",         # Kernel methods
  "nnet",            # Neural networks
  "neuralnet",       # Neural network training
  
  # Deep learning
  "torch",           # PyTorch for R
  "luz",             # High-level torch interface
  "torchvision",     # Computer vision for torch
  "tensorflow",      # TensorFlow for R
  "keras",           # High-level neural networks API
  
  # Model interpretation
  "lime",            # Local interpretable model explanations
  "DALEX",           # Descriptive machine learning explanations
  "iml",             # Interpretable machine learning
  "fastshap",        # Fast SHAP values
  "pdp",             # Partial dependence plots
  
  # Model validation
  "yardstick",       # Tidy model evaluation (part of tidymodels)
  "cvms",            # Cross-validation for model selection
  "rsample",         # Resampling methods (part of tidymodels)
  
  # Feature engineering
  "recipes",         # Preprocessing and feature engineering
  "embed",           # Embedding methods for categorical variables
  "textrecipes",     # Text preprocessing recipes
  
  # Clustering
  "cluster",         # Cluster analysis
  "factoextra",      # Cluster analysis and visualization
  "NbClust",         # Determining optimal number of clusters
  
  # Dimensionality reduction
  "umap",            # Uniform manifold approximation
  "Rtsne",           # t-SNE implementation
  "FactoMineR",      # Multivariate exploratory data analysis
  "corrr"            # Correlation analysis
))

# =============================================================================
# STATISTICAL ANALYSIS & SPECIALIZED METHODS
# =============================================================================
cat("Installing statistical analysis packages...\n")
pak::pak(c(
  # Bayesian analysis
  "brms",            # Bayesian regression models using Stan
  "rstanarm",        # Applied regression modeling via Stan
  "bayesplot",       # Bayesian model visualization
  "posterior",       # Tools for posterior analysis
  "tidybayes",       # Tidy data and geoms for Bayesian models
  
  # Survival analysis
  "survival",        # Survival analysis
  "survminer",       # Survival analysis visualization
  "flexsurv",        # Flexible parametric survival models
  
  # Mixed effects models
  "lme4",            # Linear mixed-effects models
  "nlme",            # Nonlinear mixed-effects models
  "glmmTMB",         # Generalized linear mixed models
  "broom.mixed",     # Tidy mixed-effects models
  
  # Robust statistics
  "robust",          # Robust statistics
  "robustbase",      # Basic robust statistics
  "MASS",            # Modern applied statistics
  
  # Experimental design
  "DoE.base",        # Design of experiments
  "FrF2",            # Fractional factorial designs
  "rsm"              # Response surface methodology
))

# =============================================================================
# REPORTING & REPRODUCIBLE RESEARCH
# =============================================================================
cat("Installing reporting packages...\n")
pak::pak(c(
  # Document generation
  "rmarkdown",       # Dynamic documents
  "quarto",          # Next-generation scientific publishing
  "knitr",           # Literate programming
  "bookdown",        # Authoring books and long documents
  "blogdown",        # Creating websites with R Markdown
  
  # Table formatting
  "kableExtra",      # Enhanced table formatting
  "gt",              # Grammar of tables
  "gtsummary",       # Publication-ready summary tables
  "flextable",       # Flexible table formatting
  "DT",              # Interactive data tables
  
  # Parameterized reporting
  "officer",         # Manipulate Word and PowerPoint documents
  "openxlsx",        # Excel file manipulation
  "writexl"          # Simple Excel export
))

# =============================================================================
# PERFORMANCE & PARALLEL COMPUTING
# =============================================================================
cat("Installing performance packages...\n")
pak::pak(c(
  # Parallel computing
  "parallel",        # Parallel computation (base R)
  "foreach",         # Foreach looping construct
  "doParallel",      # Parallel backend for foreach
  "future",          # Unified parallel and distributed processing
  "furrr",           # Apply purrr functions in parallel
  
  # Performance profiling
  "profvis",         # Interactive profiling
  "bench",           # High precision timing
  "microbenchmark",  # Accurate timing functions
  
  # Memory management
  "pryr",            # Memory usage and profiling tools
  "lobstr"           # Visualize R data structures
))

# =============================================================================
# DATABASE & BIG DATA
# =============================================================================
cat("Installing database and big data packages...\n")
pak::pak(c(
  # Database connections
  "DBI",             # Database interface
  "RSQLite",         # SQLite database interface
  "RPostgres",       # PostgreSQL database interface
  "odbc",            # ODBC database connectivity
  "bigrquery",       # Google BigQuery interface
  
  # Big data processing
  "arrow",           # Apache Arrow R interface
  "sparklyr",        # R interface for Apache Spark
  "disk.frame",      # Disk-based data manipulation
  
  # Data import/export
  "vroom",           # Fast reading of delimited files
  "fst",             # Fast serialization of data frames
  "qs"               # Quick serialization of R objects
))

# =============================================================================
# VERIFICATION & FINAL SETUP
# =============================================================================
cat("Installation complete! Verifying key packages...\n")

# Test loading critical packages
test_packages <- c("tidyverse", "sf", "forecast", "tidymodels", "plotly")
for(pkg in test_packages) {
  if(requireNamespace(pkg, quietly = TRUE)) {
    cat("✓", pkg, "installed successfully\n")
  } else {
    cat("✗", pkg, "installation failed\n")
  }
}

cat("\n=============================================================================\n")
cat("SETUP COMPLETE!\n")
cat("=============================================================================\n")
cat("Your R environment is now equipped with packages for:\n")
cat("• Spatial analysis and GIS\n")
cat("• Time series analysis and forecasting\n")
cat("• Advanced data visualization\n")
cat("• Machine learning and statistical modeling\n")
cat("• Reproducible research and reporting\n")
cat("• High-performance computing\n")
cat("• Database connectivity and big data processing\n")
cat("\nHappy analyzing! 🚀\n")

# Optional: Create a session info snapshot
# writeLines(capture.output(sessionInfo()), "session_info.txt")