---
title: "Weekly Viz - Metro em São Paulo"
date: "2024-10-25"
description: "Analysis of São Paulo metro ridership patterns from 2019-2023, examining the pandemic's impact and ongoing recovery across different lines, day types, and stations."
categories: ['data-visualization', 'ggplot2', 'time-series', 'transportation', 'urban-analysis']
execute:
  echo: false
  warning: false
  message: false
  fig-align: center
---

```{r setup}
library(ggplot2)
library(dplyr)
library(tidyr)
library(here)
library(ragg)
```

```{r data-import}
metro <- readr::read_csv(
  here("static/data/metro_sp.csv"),
  show_col_types = FALSE
)
```

```{r data-cleaning}
metro_labels <- c(
  "1-Azul" = "azul",
  "2-Verde" = "verde",
  "3-Vermelha" = "vermelha",
  "4-Amarela" = "amarela",
  "5-Lilás" = "lilas",
  "15-Prata" = "prata"
)

metro <- metro |>
  mutate(
    metro_line = factor(
      metro_line,
      levels = metro_labels,
      labels = names(metro_labels)
    )
  )

tbl_metro <- metro |>
  filter(
    variable == "transport",
    metric == "mdu",
    year >= 2019,
    date <= as.Date("2023-06-01")
  )

tbl_metro <- tbl_metro |>
  mutate(
    value = forecast::na.interp(value),
    trend_stl = stl(ts(value, frequency = 12), s.window = 31)$time.series[,
      "trend"
    ],
    trend_ma = RcppRoll::roll_mean(value, 7, fill = NA, align = "center"),
    .by = "metro_line"
  )
```

```{r}
colors <- paletteer::paletteer_d("rcartocolor::BluGrn")

metro_colors <- c(
  "#0153a1",
  "#008061",
  "#ee3f34",
  "#fed408",
  "#7c52a1",
  "#878e91"
)
names(metro_colors) <- metro_labels

plot_caption <- "Source: METRÔ, ViaQuatro, ViaMobilidade | Graphic: @viniciusoike"

theme_series <- theme_minimal(base_family = "Avenir") +
  theme(
    axis.line = element_line(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(
      size = 14,
      family = "Georgia",
      face = "bold"
    ),
    plot.subtitle = element_text(
      size = 10,
      color = "gray40",
      margin = margin(b = 5)
    ),
    plot.margin = margin(8, 5, 8, 5),
    plot.caption = element_text(
      color = "gray70",
      size = 8,
      hjust = 0,
      margin = margin(t = 5)
    )
  )
```

The COVID-19 pandemic fundamentally disrupted urban mobility patterns worldwide. São Paulo's metro system, which serves millions of daily passengers across six lines, provides a valuable case study for understanding how public transit ridership has evolved through the pandemic and recovery periods. This analysis examines ridership data from 2019 through mid-2023 to track the system's recovery trajectory.

## Overall System Recovery

Despite gradual improvements since the depths of the pandemic, São Paulo's metro system remains well below pre-pandemic ridership levels.

```{r}
metro_total <- metro |>
  filter(
    variable == "transport",
    metric == "total",
    year >= 2019,
    date <= as.Date("2023-06-01")
  ) |>
  summarise(
    total = sum(value, na.rm = TRUE),
    .by = "date"
  )

# Index on average 2019 values
metro_total$trend_stl <- stl(
  ts(metro_total$total, frequency = 12),
  s.window = 31
)$time.series[, "trend"]
```

```{r}
metro_facet <- metro |>
  filter(
    date >= as.Date("2018-01-01"),
    date <= as.Date("2023-06-01"),
    variable == "transport",
    metric %in% c("mdu", "msa", "mdo"),
    metro_line_num != 15,
    !is.na(value)
  )


# Compute 2018 averages for each metro-system and metric
benchmark <- metro_facet |>
  # compute 2018 averages
  filter(year == 2019) |>
  summarise(
    base_avg = mean(value, na.rm = TRUE),
    .by = c("metro_line", "metric")
  )

# Join benchmark back to main data
metro_index <- metro_facet |>
  left_join(benchmark, by = c("metro_line", "metric")) |>
  # Index and smooth values
  mutate(index = (value / base_avg) * 100) |>
  mutate(
    index_smooth = stl(
      ts(index, frequency = 12),
      s.window = "periodic"
    )$time.series[, "trend"],
    .by = c("metro_line", "metric")
  )

metro_index <- metro_index |>
  mutate(
    metric_label = stringr::str_remove(metric_label, "Média dos ")
  )
```

```{r}
line4 <- readr::read_csv(here("static/data/metro_sp_line_4_stations.csv"))

viaquatro <- line4 |>
  # Remove Vila Sônia station and filter dates from 2019 onwards
  filter(
    name_station != "Vila Sônia",
    date >= as.Date("2019-01-01")
  )

viaquatro$name_station <- factor(
  viaquatro$name_station,
  levels = unique(line4$name_station)
)

# Compute 2019 averages for each station
benchmark <- viaquatro |>
  filter(year == 2019) |>
  summarise(base_avg = mean(value), .by = name_station)

# Join benchmark averages back to the main dataset and compute the index
viaquatro_facet <- viaquatro |>
  left_join(benchmark, by = "name_station") |>
  mutate(index = value / base_avg * 100) |>
  mutate(
    index_smooth = stats::stl(
      ts(index, frequency = 12),
      s.window = "periodic"
    )$time.series[, "trend"],
    .by = "name_station"
  )

```

```{r}
ggplot(metro_total, aes(x = date)) +
  geom_point(
    aes(y = total),
    color = "white",
    fill = colors[7],
    shape = 21,
    size = 2,
    alpha = 0.6
  ) +
  geom_line(
    aes(y = trend_stl),
    color = colors[7],
    lwd = 1
  ) +
  scale_y_continuous(
    breaks = seq(25, 150, 25) * 1000,
    labels = scales::label_number(big.mark = ".", scale = 1e-3)
  ) +
  labs(
    title = "Metro Ridership still Down 30% from Pre-Pandemic Levels",
    subtitle = "Total monthly entries (millions) to the São Paulo Metro system (lines 1, 2, 3, 4, 5, and 15).",
    caption = plot_caption,
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_family = "Avenir") +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(
      size = 14,
      family = "Georgia",
      face = "bold"
    ),
    plot.subtitle = element_text(
      size = 10,
      color = "gray40",
      margin = margin(b = 5)
    ),
    plot.margin = margin(8, 5, 8, 5),
    plot.caption = element_text(
      color = "gray70",
      size = 8,
      hjust = 0,
      margin = margin(t = 5)
    )
  )
```

The chart reveals a dramatic collapse in ridership during March-April 2020, when monthly entries plummeted to roughly 25 million—less than 20% of pre-pandemic levels. While the system has steadily recovered, reaching approximately 105 million monthly entries by mid-2023, this still represents a significant 30% shortfall compared to the 130-140 million entries recorded in early 2019.

## Weekday Ridership by Line

Breaking down the recovery by individual metro lines reveals how different parts of the network have been affected.

```{r}
ggplot(tbl_metro, aes(x = date, y = trend_ma, fill = metro_line)) +
  geom_area(alpha = 0.8) +
  geom_hline(yintercept = 0) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y",
    expand = expansion(0)
  ) +
  scale_y_continuous(
    expand = expansion(0),
    labels = scales::label_number(big.mark = ".", scale = 1e-3)
  ) +
  scale_fill_manual(
    name = "",
    values = c("#171796", "#008061", "#ee3f34", "#fed408", "#7c52a1", "gray50")
  ) +
  guides(fill = "none") +
  labs(
    title = "Weekday ridership is down 20% since 2019",
    subtitle = "Average weekday ridership by metro line (millions)",
    x = NULL,
    y = NULL,
    caption = plot_caption
  ) +
  theme_minimal(base_size = 10, base_family = "Helvetica") +
  theme_series
```

The stacked area chart shows that weekday ridership patterns closely mirror the overall system trend, with all lines experiencing similar proportional declines. Lines 1 (Blue), 3 (Red), and 5 (Lilac) represent the largest shares of weekday ridership, reflecting their role as major commuter corridors.

## Recovery Patterns Across Lines and Day Types

Examining ridership indexed to 2019 levels reveals important differences in how different lines and day types have recovered.

```{r}
ggplot(metro_index, aes(date, index_smooth, color = metro_line)) +
  geom_hline(yintercept = 100) +
  geom_line(lwd = 0.75) +
  facet_grid(
    rows = vars(metric_label),
    cols = vars(metro_line)
  ) +
  scale_y_continuous(breaks = seq(20, 120, by = 20)) +
  scale_color_manual(
    values = unname(metro_colors)
  ) +
  guides(color = "none") +
  labs(
    title = "No lines have yet returned to pre-pandemic ridership levels",
    subtitle = "Monthly average ridership index by type of day (2019 = 100), smoothed values.",
    x = NULL,
    y = "Index (2019 = 100)",
    caption = plot_caption
  ) +
  theme_series +
  theme(
    strip.background = element_rect(fill = colors[7]),
    strip.text = element_text(
      color = "white",
      size = 10,
      family = "Helvetica",
      face = "bold"
    ),
    panel.background = element_rect(color = "gray10"),
    plot.title.position = "plot",
    plot.caption.position = "plot"
  )
```

A clear pattern emerges: weekend ridership (Saturdays and Sundays) has recovered more strongly than weekday commuting patterns, with several lines approaching or even exceeding 2019 levels on Sundays. This suggests that while recreational and leisure travel has rebounded, traditional work commutes remain depressed—likely due to the persistence of remote and hybrid work arrangements. Line 5 (Lilac) shows particularly volatile weekend patterns, potentially reflecting its service area's demographic and land use characteristics.

## Station-Level Analysis: Line 4 (Yellow)

Zooming into individual stations on Line 4 reveals spatial heterogeneity in the recovery process.

```{r}
#| out-width: 100%
#| fig-width: 9
#| fig-heigth: 8
ggplot(viaquatro_facet, aes(date, index_smooth)) +
  geom_point(
    aes(y = index),
    color = colors[7],
    alpha = 0.3,
    size = 0.5
  ) +
  geom_line(color = colors[7], size = 0.8) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "gray40") +
  scale_y_continuous(breaks = seq(0, 120, by = 20)) +
  labs(
    title = "Station-Level Recovery Varies Across Line 4",
    subtitle = "Monthly ridership index by station on the Yellow Line (2019 = 100)",
    x = NULL,
    y = "Index (2019=100)",
    caption = "Source: ViaQuatro | Graphic: @viniciusoike"
  ) +
  facet_wrap(vars(name_station), ncol = 5) +
  theme_series
```

Station-level data reveals significant variation in recovery trajectories. Some stations, like Oscar Freire and Paulista—located in São Paulo's high-end commercial and business districts—show stronger recovery, potentially reflecting their role as destinations for both work and leisure activities. In contrast, primarily residential stations show more modest recovery, consistent with reduced commuting demand. Notably, several stations have surpassed their 2019 ridership levels by 2023, suggesting localized growth or shifting travel patterns within the metro network.

## Implications

The persistent gap between current and pre-pandemic ridership levels reflects fundamental changes in urban mobility behavior. The stronger recovery in weekend versus weekday ridership indicates that remote and hybrid work arrangements have become structural features of São Paulo's labor market, not temporary pandemic responses. This shift has significant implications for transit planning, fare revenue projections, and the broader urban economy that depends on commuter spending in business districts.
